RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-12-29 manuscripts/manuscript.Rmd
Rewrite/arrangemnt of introduction.
This is to tighten up the text for submission as an Application Note.

The popular statistical and mathematical computing platform R [@R] arguably provides the most extensive and widely used collection of
comparative phylogenetic methods tools <!-- needs a citation to support it -->.
Yet, R users lack a convenient and extensible way to import, exchange and archive 
phylogenetic data, trait data, and their accompanying rich metadata in
a data exchange standard that comprehensively supports current data
exchange needs, is predictably machine-readable, and is forward compatible.
In response to the rapid increase in the variety of data
relevant to comparative analysis, and the rising data interoperability
needs of integrative comparative research approaches, such an exhcange standard,
called NeXML, has recently
been developed [@Vos_2012], and is beginning to be adopted. One of the
hurdles to broader adoption of any exchange standard is the
availability of easy-to-use programmtic interfaces (APIs) to the
format, in particular for developers of tools that consume or generate
relevant data, so that supporting the standard for reading input and
writing output data does not require additional effort, compared to,
say, using other but inferior exchange formats. Here we present such
an API in the form of an R package, called RNeXML. RNeXML aims to
provide easy programmatic access to reading and writing NeXML documents,
tailored for the kinds of use-cases that will be common for users and
developers of the wealth of evolutionary, and in particular
comparative analysis methods within the R ecosystem.

Increasingly many comparative research questions have come to exploit
the opportunities from, and thus rely on the ability to reuse and
integrate previously generated or published data and
metadata. However, the ever expanding needs of the evolutionary
research community are rapidly outpacing the capabilities of the
currently still most widely used exchange standards, which were all
developed a decade or more ago. This has resulted in a radiation of
different data representations and exchange standard "flavors" that
are no longer interoperable at the very time when the growth of
available data and methods has made that interoperability most
valuable. The problem is particularly acute in light of the challenges
for finding and sharing phylogenetic data
[@Drew_2013; @Stoltzfus_2012], in particular in a way that is not
subject to loss of most data and metadata semantics, or that is
amenable to predictable machine interpretation [@Cranston_2014]. For
example, the currently popular NEXUS file format [@Maddison_1997] cannot consistently
represent horizontal gene transfer or ambiguity in reading a character
(such as a DNA sequence base pair).  This has led to
modifications of NEXUS in different ways for different needs, with the unfortunate
result that NEXUS files generated by one program can be incompatible
with another [@Vos_2012]. Without a formal grammar, software based on 
NEXUS files may also make inconsistent assumptions about tokens, quoting,
or element lengths.  @Vos_2012 estimates that as many as 15% of the NEXUS
files in the CIPRES portal contain unrecoverable but hard to diagnose errors.

A detailed account of how the NeXML standard addresses these and other
relevant challenges can be found in @Vos_2012. In brief, NeXML was
designed with the following important properties. First, NeXML is defined
by a precise grammar that can be programmatically **validated**. This
ensures that any NeXML file can be expected to be parsed and read
without errors by any NeXML-compliant software, provided the file
passes validation. Second, NeXML is **extensible**: a user can define
representations of new, previously unanticipated information (as we
will illustrate) without violating its defining grammar. Third and
most importantly, NeXML is **semantically rich**.
The use of formal semantics in NeXML makes its data representation
not only machine-readable in a predictable manner, but also enables machine reasoning about
the data and metadata (see @Parr2011 for an overview on semantic reasoning in 
evolutionary informatics). For instance, through the use of semantics we can
expect the computer to reason about taxonomic concepts
(as we illustrate in the supplement). 
<!-- Sorry, just trying to contrast what it means to make a 'semantic' query 
vs a non-semantic one.  Most R users won't be familiar with SQL queries either anyway.  Let me know if this is better.-->

The R language and platform [@R] has become a powerful ecosystem for
evolutionary analysis, with over 30 packages available alone for
comparative phylogenetics, several of which have accumulated over 
1000 citations [@taskview]. `RNeXML` makes the sizable amount of
richly annotated phylogenetic data already existing in NeXML format
available to R users and R's ecosystem, in a way, as we show here,
that is no more involved than reading or writing data from other
phylogenetic data formats. `RNeXML` also carries immediate benefits
for its users. For example, comparative analysis R packages and users
frequently add their own metadata annotations to the phylogenies they work with, including annotations of species,
stochastic character maps, trait values, model estimates and parameter 
values. `RNeXML` affords R the capability to harness machine-readable
semantics and an extensible metadata schema to capture, preserve, and
share these and other kinds of information, all through an API instead
of having to understand in detail the schema underlying the NeXML
standard. To assist users in meeting the rising bar for best practices
in data sharing in phylogenetic research [@Cranston_2014], `RNeXML`
captures metadata information from the R environment to the extent
possible, and applies reasonable defaults. Beyond these more immediate
benefits, `RNeXML` plugs the R ecosystem into the growing informatics
infrastructure of evolutionary informatics [@Parr2011], in which
interoperable data and machine-readable semantics play important
enabling roles.

The rest of this paper is organized as follows. First, we introduce
the basic use and concepts of RNeXML for reading and writing
phylogenetic trees and character trait data from and to NeXML. We then
focus on RNeXML's support for reading and writing the kind of
semantically rich metadata that most sets NeXML apart from
other phylogenetic data formats. We include examples of
increasing complexity to illustrate concepts and tools from biodiversity
informatics that allow users to express and interpret complicated data
structures automatically.  We end this section with a concrete example illustrating
how RNeXML facilitates harnessing the forward-compatibility property
of NeXML, by devising a serialization for stochastic character
mappings, a data annotation generated by existing R packages but
lacking a standard exchange format. Next we illustrate how RNeXML
promotes data archiving and sharing by providing commands for directly
publishing NeXML files to a data repository. We conclude the paper
with a brief description of how the software has been
implemented, what quality control measures are in place, as well as
information about software reuse and getting support.

This paper has been written as a dynamic document in R Markddown
[@Xie_2013], which directly exercises all of the code blocks shown every time
the typeset version is regenerated, providing immediate feedback on
whether the code as shown is complete and continues to work. It also
means that the reader may paste the code blocks into their own R
terminal to follow along with the examples.

# Introduction to RNeXML

Conceptually, a NeXML document has the following components: (1) phylogeny
data, (2) character or trait data in matrix form, (3) operational
taxonomic units (OTUs), and (4) metadata. When reading and writing NeXML,
`RNeXML` aims to map these to and from, respectively, their most
natural representations in R. These are the `phylo` object from the
`ape` package [@Paradis_2004], R's native `data.frame` list structure for
phylogenies, character matrices, and metadata, respectively. Note that
unlike NeXML or NEXUS, the `phylo` object stores taxonomic information
only as tip labels and not as a separate block of data.

NeXML allows attaching metadata to any data element, and even to
metadata themselves. At the level of the document as a whole or the
phylogeny, metadata can provide bibliographic and provenance
information, for example about the study as part of which the
phylogeny was generated or applied, and with which methods it was
generated. In the form of annotations, metadata can also be attached
to very specific elements of the data, such as individual OTUs, nodes,
or even edges of the phylogeny.

## Getting Started

As for other R packages, prior to using `RNeXML`, the package needs to
be installed and loaded. The current stable version can be installed from
the CRAN library[^1]:

>>>>

Users of the popular statistical and mathematical computing platform R
[@R] enjoy a wealth of readily installable comparative phylogenetic
methods and tools <!-- needs a citation to support it -->. Exploiting
the opportunities arising from this wealth for complex and integrative
comparative research questions relies on the ability to reuse and
integrate previously generated or published data and metadata. The
resulting expanding data exchange needs of the evolutionary research
community are rapidly outpacing the capabilities of the currently
still most widely used data exchange standards, which were all
developed a decade or more ago. This has resulted in a radiation of
different data representations and exchange standard "flavors" that
are no longer interoperable at the very time when the growth of
available data and methods has made that interoperability most
valuable. In response to the unmet needs for standardized data
exchange in phylogenetics, a modern XML-based exchange standard,
called NeXML, has recently been developed [@Vos_2012], and is
beginning to be adopted. NeXML comprehensively supports current data
exchange needs, is predictably machine-readable, and is forward
compatible. To make the capibilities of NeXML available to R users in
an easy-to-use form, and to lower the hurdles to adoption of the
standard, we present RNeXML, an R package that aims to provide easy
programmatic access to reading and writing NeXML documents, tailored
for the kinds of use-cases that will be common for users and
developers of the wealth of evolutionary, and in particular
comparative analysis methods within the R ecosystem.

The exchange problem for phylogenetic data is particularly acute in
light of the challenges in finding and sharing phylogenetic data
without the otherwise common loss of most data and metadata semantics
[@Drew_2013; @Stoltzfus_2012], or with semantics that is predictably
machine interpretable [@Cranston_2014]. For example, the still popular
NEXUS file format [@Maddison_1997] cannot consistently represent
horizontal gene transfer or ambiguity in reading a character (such as
a DNA sequence base pair).  This has led to modifications of NEXUS in
different ways for different needs, with the unfortunate result that
NEXUS files generated by one program can be incompatible with another
[@Vos_2012]. Without a formal grammar, software based on NEXUS files
may also make inconsistent assumptions about tokens, quoting, or
element lengths.  @Vos_2012 estimates that as many as 15% of the NEXUS
files in the CIPRES portal contain unrecoverable but hard to diagnose
errors.

A detailed account of how the NeXML standard addresses these and other
relevant challenges can be found in @Vos_2012. In brief, NeXML was
designed with the following important properties. First, NeXML is
defined by a precise grammar that can be programmatically
**validated**. This ensures that any NeXML file can be expected to be
parsed and read without errors by any NeXML-compliant software,
provided the file passes validation. Second, NeXML is **extensible**:
a user can define representations of new, previously unanticipated
information (as we will illustrate) without violating its defining
grammar. Third and most importantly, NeXML is **rich in computable
semantics**. The formal semantics used in NeXML is machine-readable in
a predictable manner, and also enables machine reasoning about the
data and metadata (see @Parr2011), especially when metadata attributes
and values are anchored in appropriately axiomatized ontologies.

`RNeXML` not only allows R's ecosystem to read and write data in
the NeXML format through an interface that is no more involved than
reading or writing data from other phylogenetic data formats, but also
carries immediate benefits for its users. For example, comparative
analysis R packages and users frequently add their own metadata
annotations to the phylogenies they work with, such as annotations
of species, stochastic character maps, trait values, model estimates
and parameter values. `RNeXML` affords R the capability to harness
machine-readable semantics and an extensible metadata schema to
capture, preserve, and share these and other kinds of information, all
through an API instead of having to understand in detail the schema
underlying the NeXML standard. To assist users in meeting the rising
bar for best practices in data sharing in phylogenetic research
[@Cranston_2014], `RNeXML` captures metadata information from the R
environment to the extent possible, and applies reasonable
defaults.

# The RNeXML package

`RNeXML` is an R package entirely written in R, and the current stable
version can be installed from the CRAN library[^1]:
